---
title: "Class 05: Data Visualization with GGPLOT"
author: "Dahlia Loomis"
date: 04/19/2023
format: pdf
---

# Base R Plotting

We are going to start by generating the plot from Class 04. This code plots the car `dataset.`

```{r}
plot(cars)
```

# Ggplot2

First, we need to install the package. We do this by using the `install.packages` command.

```{r}
#install.packages('ggplot2')
```

After that, we need to load the package.

```{r}
library(ggplot2)
```

Next, we will generate a plot with the cars dataframe by using ggplot2. The following code demonstrates the components and purpose of the different functions that allow us to build up to the actual plot.

```{r}
ggplot(data=cars)
ggplot(data=cars) + 
  aes(x=speed, y=dist)
ggplot(data=cars) + 
  aes(x=speed, y=dist) +
  geom_point()
ggplot(data=cars) + 
  aes(x=speed, y=dist) +
  geom_point() +
  geom_smooth(method = 'lm') +
  labs(title= 'Distance vs. Speed',
       subtitle = 'A nice subtitle!', 
       caption= 'BIMM 143', 
       x = 'Speed (MPH)', 
       y = 'Distance (feet)') +
  theme_bw()

```

# Plotting Gene Expression Data

Loading data for gene expression

```{r}
#load data
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

Initial ggplot

```{r}
ggplot(data = genes) +
  aes(x = Condition1, y = Condition2) +
  geom_point()
  
```

Add color to the plot

```{r}
p1 <- ggplot(data = genes) +
  aes(x=Condition1, y=Condition2, col=State) +
  geom_point()
```

Let's change the color scheme

```{r}
p2 = p1 + 
  scale_color_manual(values = c("blue", "gray", "red")) 
```

Let's add some labels.

```{r}
p2 + labs(title = 'Differential Gene Expression', x = 'Control', y = 'Drug Treatment', caption = 'BIMM 143 - Class 05')
```

# Questions

**Q1.**

For which phases is data visualization important in our scientific workflows?

**For explanatory data analysis (EDA), communication of results, and detection of outliers.**

True or False? The ggplot2 package comes already installed with R?Plotting Gene Expression Data

**FALSE**

**Q2.**

Which plot types are typically NOT used to compare distributions of numeric variables?

**Network graphs are typically not used to compare distributions of numeric variables.**

Which statement about data visualization with ggplot2 is incorrect?

**Ggplot2 is the only way to create plots in R**

**Q3.**

Which geometric layer should be used to create scatter plots in ggplot2?

**geom_point**

**Q4.**

```{r}
nrow(genes)
ncol(genes)
colnames(genes)

table(genes[,'State'])
```

Use the `nrow()` function to find out how many genes are in this dataset. What is your answer?

**5196**

Use the `colnames()` function and the `ncol()` function on the `genes` data frame to find out what the column names are (we will need these later) and how many columns there are. How many columns did you find?

**4**

Use the `table()` function on the `State` column of this data.frame to find out how many 'up' regulated genes there are. What is your answer?

**There are 127 'up' regulated genes**

Using your values above and 2 significant figures. What fraction of total genes is up-regulated in this dataset?

```{r}
table(genes[,'State'])/ nrow(genes)*100


```

**2.44**
